A

### 2015-10-20 paper-draft/paper.tex
updates to thesis

Section \ref{sec:preliminaries}
lists related works,
describes the cyclic reduction algorithm
and the specific issues in its implementation on GPUs.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

@@

@@

\begin{align} \label{eqn:forward-reduction}
& a^{\prime}_i = -a_{i-1}k_1 & \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
& c^{\prime}_i = -c_{i+1}k_2 & \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-rhs}&
\end{align}

The 2-by-2 system of equations is solved trivially,
yielding $x_n$ and $x_{n/2}$.
In the backward substitution phase,
every odd-indexed unknown $x_i$ is solved for by
substituting the known values of $x_{i-1}$ and $x_{i+1}$:

@@

@@

@@

We note from
Equations \ref{eqn:forward-reduction}-\ref{eqn:forward-reduction-rhs}
that each forward reduction step involves 12 operations.
As the LHS matrix is constant,
it is advantageous to pre-compute the coefficients appearing
in the forward reduction step.
For a general tridiagonal matrix,
in addition to the $3n$ original coefficient arrays
$a$, $b$ and $c$,
this requires extra storage of 
$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
coefficients.
This storage cost is greatly amortized
when solving multiple right-hand sides.

>>>>

Section \ref{sec:preliminaries}
describes algorithms for solving tridiagonal systems on GPUs.
The cyclic reduction algorithm in particular, is discussed in detail,
and issues with its implementation on GPUs is discussed.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

@@

\subsection{Solving tridiagonal systems on the GPU}

@@

\begin{align}
& a^{\prime}_i = -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1}& \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2}& \\
& c^{\prime}_i = -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3}& \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}&
\end{align}

The 2-by-2 system of equations is solved trivially,
yielding $x_n$ and $\frac{x_n}{2}$.
In the backward substitution phase,
every odd-indexed unknown $x_i$ is solved for by
substituting the known values of $x_{i-1}$ and $x_{i+1}$:

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/parallel-cyclic-reduction-single-path.eps}
\end{center}
\caption{Parallel cyclic reduction - one reduction path highlighted}
\label{fig:parallel-cyclic-reduction-single-path}
\end{figure}

Parallel cyclic reduction begins by
performing the forward reduction step on
both the odd and even indexed equations.
This results in two systems of size $\frac{n}{2}$ each.
Forward reduction is applied again to each of these systems
resulting in four systems of size $\frac{n}{4}$ each.
The process is repeated till $\frac{n}{2}$ 2-by-2 systems are produced,
which can each be solved trivially.
Figure \ref{fig:parallel-cyclic-reduction-single-path}
shows PCR with one of these ``reduction paths'' highlighted.

Our proposed solver is a specialization of cyclic reduction,
so we discuss the pertinent implementation issues on GPUs.

\subsection{GPU architecture}

Writing applications that effectively exploit the GPU
requires a thorough understanding of the underlying architecure.
We provide the pertinent details here,
and refer to \cite{GPUcomputingera} for a more complete picture.
Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
built on the ``Kepler'' architecture,
an overview of which is available at \cite{Keplerwhitepaper}.

Applications that use the GPU include special
pieces of code that are executed on the GPU,
referred to as \emph{kernels}.
In C, for example, kernels are written as functions,
and are called with (almost) the same conventions.
A kernel is executed in parallel by several threads,
which are organized into a \emph{grid} of \emph{thread blocks}
(or just \emph{blocks}).
As a requirement, thread blocks must be able to run
independent of each other,
in series or in parallel.

From the hardware perspective, an NVIDIA GPU may be viewed primarily as
a collection of so-called Streaming Microprocessors (SMs).
When a kernel is launched (with a specified number of thread blocks),
each block is assigned to an SM.
Threads within a thread block can execute concurrently on the SM,
and an SM can execute several thread blocks concurrently.
The number of thread blocks that an SM can execute concurrently
is limited by a number of factors,
and maximizing this number is often key to obtaining good performance.

Every thread in every block has access to so-called \emph{global memory},
which can also be accessed by the CPU (or \emph{host}) via the PCI-e bus.
Further,
threads within a thread block have access
to a common, fast, limited \emph{shared memory}.
The contents of shared memory are managed by the kernel code,
and shared memory is often viewed as an explicitly controlled cache.
Each thread also has private \emph{local memory},
and access to extremely fast registers.
Unlike CPUs, the GPU has a large number of registers---a
thread executing a kernel will
typically attempt to store the variables defined in registers,
before spilling over to local memory, which is much slower.

Thread access to global memory is slow,
and is especially inefficient when
successive threads in a block
access locations that are far apart in memory---known
as \emph{uncoalesced} memory access.
Shared memory access can be expected to be
much faster than global memory---however,
the amount of shared memory per SM is limited
(48 KiB for current GPUs).
The amount of shared memory actually used by each block,
can thus limit the number of blocks that concurrently run on each SM.
The number of registers per SM is also limited,
and the use of registers can similarly limit
the number of concucrrently running blocks.
Finally, memory transfers between the CPU (host) and GPU (device)
are extremely expensive, limited by the bandwidth of the PCI-e bus.

\subsection{Cyclic reduction on GPU}

@@

We note that solving the equations
\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
for each grid line and at each time step of the simulation
is wasteful.
It may be benefitial to
\emph{precompute} and store the coefficients appearing at
each step of the forward reduction,
and to load these values from memory when required.

\subsection{General tridiagonal matrix}

For a general tridiagonal matrix,
in addition to the $3n$ original coefficient arrays
$a$, $b$ and $c$,
this requires extra storage of 
$\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...=n$
elements each
for $a_i$, $b_i$, $c_i$, $k1$ and $k2$
appearing at the forward reduction steps.
This additional storage cost is greatly amortized
when solving the same system for multiple right-hand sides.
But the additional arrays
still need to be loaded into shared memory for each thread block.
This raises the shared memory storage from $4n$ to $9n$,
which significantly impacts the number of
concurrently running blocks.

\subsection{Near-Toeplitz matrix}

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/cyclic-reduction-boundaries.eps}
\end{center}
\caption{Indices affected by boundaries in cyclic reduction}
\label{fig:cyclic-reduction-boundaries}
\end{figure}

Precomputing the forward reduction coefficients
becomes especially effective for near-Toeplitz matrices.
At each reduction step,
the coefficients $a_i$, $b_i$, $c_i$, $k1_i$ and $k2_i$ 
are the same for all indices $i$,
except for indices ``polluted'' by the boundary conditions
(Figure \ref{fig:cyclic-reduction-boundaries}).
Ignoring the boundaries, then, we find that the precomputed coefficients
require $5(log_2(n) - 1)$ storage---far lower than the $5n$ required for the
general case.
Further, a careful analysis of the equations
\ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
reveals that the following extra arrays are required
to store the boundary information:

\begin{enumerate}
    \item The values of $b_1$ at each reduction step
    \item The values of $k_{1,1}$ at each reduction step
    \item The values of $k_{1,n}$ at each reduction step
    \item The values of $a_n$ and $b_n$ at the final reduction step
        (required for the 2-by-2 solve)
\end{enumerate}

This puts the total storage for precomputed coefficients
at $8(log_2(n) - 1)+2$.

@@

We note from
Equations \ref{eqn:forward-reduction-1}-\ref{eqn:forward-reduction-3}
that each forward reduction step involves 12 operations.
As the LHS matrix is constant,
it is advantageous to pre-compute the coefficients appearing
in the forward reduction step.
For a general tridiagonal matrix,
in addition to the $3n$ original coefficient arrays
$a$, $b$ and $c$,
this requires extra storage of 
$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
coefficients.
This storage cost is greatly amortized
when solving multiple right-hand sides.
